The XAI Paradox: Systems that Perform Well for the Wrong Reasons

### 2018-08-28 Paper.tex
Small changes

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Successful machine learning approaches are often regarded as ``black box'' systems: they perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that can provide explanations underlying their internal reasoning. 
In this study, we investigate to
what extent systems learn explanatory structures that make sense in
a domain. 
Using artificial datasets whose internal structure is determined beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. 
We generate datasets from predetermined formal conditions defined in terms of symmetric Boolean functions. When using datasets generated by conjunctions of conditions, we find that trained systems perform well on the conjunction problem, but---paradoxically---not on the individual conditions separately. This implies that a confounding
%Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. In particular, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding 
structure within the data is learnt that still allows making the correct decisions. In other words, the systems perform well, but for the wrong reasons, providing a puzzle for the XAI movement.
%With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
%Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

@@

For any pair of conditions (Parity, Exact; Parity, M-out-of-N; M-out-of-N, Exact), the following datasets are generated: a training set, a general test set and a specific test set for each condition. First of all, a training set consists of 150,000 instances where half of the instances are generated randomly such that their output value is true, and the other half are generated randomly such that their output value is false.  In the latter half, the instances are generated such that they fail on each of its conditions equally. With two conditions for example, this means that the latter half (the half where the output is false) would consists for 50\% out of instances where the first condition is false and for 50\% out of instances where the second condition is false. If a condition is not specifically set up to fail, its variables are provided with random values. The general test set consists of 150,000 instances, and is generated in the exact same fashion.

For each condition, also a specific test set is generated, used to determine how well a system is able to learn that specific condition. Given such a condition, its test set is constructed by generating all of the possible input values that the condition can have. For example, the XOR condition (or parity function with $n$=2) has $2^2=4$ possible values and would therefore have a specific test set with 4 instances. The variables of any additional conditions are given random values such that these conditions are satisfied. This ensures that the output is dependent on only the single condition for which the test set is generated. %This allows us to measure how well that particular condition is learned by the system.

@@

The accuracies on the general test set (Figure \ref{fig:Heatmap_ME_DT_general}) show the overall performance of the decision tree system on the combined problem. Unsurprisingly, more M-out-of-N and exact value function variables lead to a decrease in accuracy; as the problem becomes more complex, it is more difficult for the system to learn. Figure \ref{fig:Heatmap_ME_DT_M} displays how well the M-out-of-N function is learned by the system, and a similar decrease in accuracy can be observed when more variables are used in the data. Interestingly, the number of M-out-of-N variables has a bigger influence on the accuracy on this test set than the number of exact value variables. This could be because functions with more variables are more difficult for the systems to learn. In Figure \ref{fig:Heatmap_ME_DT_E}, which displays how well the exact value function is learned, a different effect can be observed. The accuracy on this test set decreases with more exact value variables, but increases with more M-out-of-N variables. So a larger number of
exact value variables decreases how well the M-out-of-N function is learned, while a larger number of M-out-of-N variables increases how well the exact function is learned.

@@

\section{Discussion}
Our findings show that in our generalized setting with symmetric Boolean functions a high performance in terms of classification accuracy does not guarantee that the system has learned the conditions that define the dataset, as evident by Tables \ref{tbl:interactions_DT} and \ref{tbl:interactions_NN}. Also we saw that when dealing with interactions, the number of variables of a function can either increase or decrease how well a system learns the function, depending on the other function that is in the dataset. The accuracy on the parity test set increases with a low number of parity variables in both the interaction with the exact value function and the M-out-of-N function. The accuracy on the exact value test set increases with a high number of exact variables in the interaction with the parity function and with a low number of exact variables in the interaction with the M-out-of-N function. Similarly, the accuracy on the M-out-of-N test set increases with a high number of M-out-of-N variables in the interaction with the parity function, and with a low number of M-out-of-N variables in the interaction with the exact value function. Table \ref{tbl:highlow}  shows that higher numbers of parity variables always have a negative effect on the how well the parity function is learned, regardless of other the functions that define the dataset. This is also true for the other two functions, except when interacting with the parity function; when interacting with the parity function, a high number of exact/M-out-of-N variables has a positive effect. Earlier results %%%%BV REFERENCE?
already showed that the parity function is more difficult to learn than the other two functions. This, combined with the finding of Table \ref{tbl:highlow} suggests that the interference of the parity function has a greater negative impact on how well the the exact and M-out-of-N function are learned than the number of variables of said function. This could explain why the increase in variables creates a increase in learning the exact and M-out-of-N functions in combination with the parity function.

@@

\section{Conclusion}
\noindent In the future, machine learning algorithms that simply execute a task with a high performance will not suffice; an explanation of their decision making will be required \cite{gunning2017explainable}. This explanation, however, is dependent on the reasoning of the systems; if the reasoning is unsound, the explanation will be unsound as well. This study has shown that machine learning algorithms do not always internalize the structure of their training data as we would expect. With regards to the emerging XAI techniques and the explainable models that they generate, such unsound rationales underlying machine learning systems can form a hindrance in creating an understandable explanation.

>>>>

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Successful machine learning approaches are often regarded as ``black box'' systems: they perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that can provide explanations underlying their internal reasoning. In this study, we investigate to what extent systems learn explanatory structures that make sense in a domain. Using artificial datasets whose internal structure is determined beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. We generate datasets whose outputs are based on predetermined formal input conditions defined in terms of symmetric Boolean functions. When the output of the generated datasets are dependent on a conjunctions of input conditions, we find that trained systems perform well on the conjunction problem, but---paradoxically---not on the individual conditions separately. This implies that a confounding
%Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. In particular, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding 
structure within the data is learnt that still allows making the correct decisions. In other words, the systems perform well, but for the wrong reasons, providing a puzzle for the XAI movement.
%With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
%Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

@@

For any pair of conditions (Parity, Exact; Parity, M-out-of-N; M-out-of-N, Exact) and all possible combinations of parameters $m$ and $n$, the following datasets are generated: a training set, a general test set and a specific test set for each condition. First of all, a training set consists of 150,000 instances where half of the instances are generated randomly such that their output value is true, and the other half are generated randomly such that their output value is false.  In the latter half, the instances are generated such that they fail on each of its conditions equally. With two conditions for example, this means that the latter half (the half where the output is false) would consists for 50\% out of instances where the first condition is false and for 50\% out of instances where the second condition is false. If a condition is not specifically set up to fail, its variables are provided with random values. The general test set consists of 150,000 instances, and is generated in the exact same fashion.

For each condition, a specific test set is also generated, which is used to determine how well a system is able to learn that specific condition. Given such a condition, its test set is constructed by generating all of the possible input values that the condition can have. For example, the XOR condition (or parity function with $n$=2) has $2^2=4$ possible values and would therefore have a specific test set with 4 instances. The variables of any additional conditions are given random values such that these conditions are satisfied. This ensures that the output is dependent on only the single condition for which the test set is generated. %This allows us to measure how well that particular condition is learned by the system.

@@

The accuracies on the general test set (Figure \ref{fig:Heatmap_ME_DT_general}) show the overall performance of the decision tree system on the combined problem. Unsurprisingly, more M-out-of-N and exact value function variables lead to a decrease in accuracy; as the problem becomes more complex, it is more difficult for the system to learn. Figure \ref{fig:Heatmap_ME_DT_M} displays how well the M-out-of-N function is learned by the system, and a similar decrease in accuracy can be observed when more variables are used in the data. Interestingly, the number of M-out-of-N variables has a bigger influence on the accuracy on this test set than the number of exact value variables. This could be because functions with more variables are more difficult for the systems to learn. In Figure \ref{fig:Heatmap_ME_DT_E}, which displays how well the exact value function is learned, a different effect can be observed. The accuracy on this test set decreases with more exact value variables, but increases with more M-out-of-N variables. So a larger number of
exact value variables decreases how well the M-out-of-N function is learned, while a larger number of M-out-of-N variables increases how well the exact function is learned. The accuracies of the neural network (Figure \ref{fig:Heatmap_NN_ME} show similar, but less pronounced trends.

@@

\section{Discussion}
Our findings show that in our generalized setting with symmetric Boolean functions a high performance in terms of classification accuracy does not guarantee that the system has learned the conditions that define the dataset, as evident by Tables \ref{tbl:interactions_DT} and \ref{tbl:interactions_NN}. Also we saw that when dealing with interactions, the number of variables of a function can either increase or decrease how well a system learns the function, depending on the other function that is in the dataset. The accuracy on the parity test set increases with a low number of parity variables in both the interaction with the exact value function and the M-out-of-N function. The accuracy on the exact value test set increases with a high number of exact variables in the interaction with the parity function and with a low number of exact variables in the interaction with the M-out-of-N function. Similarly, the accuracy on the M-out-of-N test set increases with a high number of M-out-of-N variables in the interaction with the parity function, and with a low number of M-out-of-N variables in the interaction with the exact value function. Table \ref{tbl:highlow}  shows that higher numbers of parity variables always have a negative effect on the how well the parity function is learned, regardless of other the functions that define the dataset. This is also true for the other two functions, except when interacting with the parity function; when interacting with the parity function, a high number of exact/M-out-of-N variables has a positive effect. This suggests that the interference of the parity function has a greater negative impact on how well the the exact and M-out-of-N function are learned than the number of variables of said function. This could explain why the increase in variables creates a increase in learning the exact and M-out-of-N functions in combination with the parity function.

@@

\section{Conclusion}
\noindent In the future, machine learning algorithms that simply execute a task with a high performance will not always suffice; an explanation of their decision making will be required \cite{gunning2017explainable}. This explanation, however, is dependent on the reasoning of the systems; if the reasoning is unsound, the explanation will be unsound as well. This study has shown that machine learning algorithms do not always internalize the structure of their training data as we would expect. With regards to the emerging XAI techniques and the explainable models that they generate, such unsound rationales underlying machine learning systems can form a hindrance in creating an understandable explanation.
